---
title: "Life expectancy evolution in France"
subtitle: "⚔<br/>since 1816"
author: "Corentin Guillaumet, Skandar Mabrouk & Lucas Veeravalli"
institute: "University of Paris"
date: "December 14th 2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:9
---
class: center, middle
#Summary
- Evolution of mortality quotient since 1816
- How to obtain an estimation for the future
- Estimating life expectancies

---
class: center, top
##I. Evolution of mortality quotient since 1816
```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
if(! 'pacman' %in% installed.packages()[,1]){
  install.packages("pacman")
}
pacman::p_load(tidyverse)
pacman::p_load(shiny)
pacman::p_load(knitr)
pacman::p_load(png)
pacman::p_load(plotly)
pacman::p_load(readr)
pacman::p_load(StMoMo)
pacman::p_load(ggfortify)
pacman::p_load(pathlibr)
pacman::p_load(demography)
pacman::p_load(glue)

fpath <- 'full_life_table.Rds'  # once you have downloaded the file
if (! file.exists(fpath)){
  cat(glue('{fpath} should be in working directory!'))
} else {
  life_table <- readr::read_rds(fpath)
  glimpse(life_table)
}


```

```{r, echo=FALSE, include=FALSE}
pivot_life <- function(life_table) {
life_table_pivot <- life_table %>%
  select(Country, Gender, Year, Age, qx) %>%
    pivot_wider(
      id_cols = c("Country","Gender","Year"),
      names_from = Age,
      values_from = qx
    )
return (life_table_pivot)
}
life_table_pivot <-pivot_life(life_table)
  
```
.middle[
.pull-left[
```{r}

plot1 <- life_table_fr_evol <- life_table %>% filter(Country =='France', Gender != "Both") %>%
  ggplot() +
  geom_line(mapping = aes(y= qx, col = Gender, x= Age)) +
  aes(frame = Year, text = qx) +
  theme(axis.text.x=element_text(angle=-45, hjust=0, vjust=1)) +xlab("Age") +ylab("Mortality quotient")+
  theme(plot.title=element_text(size=12, hjust=0.5, face="bold", colour="maroon", vjust=-1)) +
  theme(plot.subtitle=element_text(size=10, hjust=0.5, face="italic", color="black"))

```
]
.pull-right[
```{r, echo=FALSE}
ggplotly(plot1, height = 400, width=550)
```
]
]

---
class: center, middle
##II. A. How to obtain an estimation for the future (Computing Lee Carter model)
.pull-left[
```{r, echo = TRUE}
#FEMALE
Years <- 1933:1995
T <- length(Years)
Age <- 1:110
A_FRF <- life_table_pivot %>%
  filter(Country == "France", Gender == 'Female', Year %in% Years) %>%
  select(-Gender, -Country, -Year) %>%
  log() %>%
  as.matrix()
M_FRF<- (diag(1,T) - ((1/T)*matrix(1,T,1) %*% t(matrix(1,T,1)))) %*% A_FRF
svdFRF <- svd(M_FRF,1,1)
sFRF <- svdFRF$d[1]
uFRF <- -1 *(svdFRF$u*sFRF)
vFRF <- -1 *svdFRF$v
axFRF <- colMeans(A_FRF)
```
]
.pull-right[
```{r, echo=FALSE}
#plot(Age, ax)
#plot(Age, v, ylab = "Bx")
#plot(Years, u, ylab = "K")
data_FRF <- tibble(age = 0:109, axFRF)
#alpha
g_1_FRF <- ggplot(data_FRF) + geom_point(aes(age,axFRF)) +
  geom_line(aes(age,axFRF), col = "red") +
  theme_bw() +
  ggtitle("Alpha estimating of France Female") +
  labs(y = bquote(hat(alpha)[x]^"(1)"))
#beta
g_2_FRF <- ggplot(data_FRF) + geom_point(aes(age,vFRF)) +
  geom_line(aes(age,vFRF), col = "red") +
  theme_bw() +
  ggtitle("Beta estimating of France Female") +
  labs(y = bquote(hat(beta)[x]^"(1)"))

#kappa
data_period_FRF <- tibble(year = 1933:1995, uFRF)

g_3_FRF <- ggplot(data_period_FRF) + geom_point(aes(year,uFRF)) +
  geom_line(aes(year,uFRF), col = "red") +
  theme_bw() +
  ggtitle("kappa estimating of France Female") +
  labs(y = bquote(hat(kappa)[t]^"(1)"))
gridExtra::grid.arrange(g_1_FRF,g_2_FRF,g_3_FRF, ncol = 2)
```
]
---
class: center, middle
###II. B. Comparing Norms (SVD Rank 2 vs Lee Carter)
.pull-left[
```{r}
res_FRF <- svd(A_FRF)
matFRF <- (res_FRF$u[,1:2]) %*% diag(res_FRF$d[1:2])%*%t(res_FRF$v[,1:2])

matLEEFRF = matrix(0,nrow = 63 , ncol = 110)
for (i in 1:110){
  matLEEFRF[,i] = t(axFRF)[i] +vFRF[i]*uFRF
}

```
]
.pull-right[
```{r}
norm(matFRF - as.matrix(A_FRF) , "F") #9.684005
norm(matLEEFRF -as.matrix(A_FRF), "F") #10.07679
```
]
---
class: center, middle
##II. C. Computing and displaying Lee Carter parameters
.middle[
```{r,echo=FALSE,include=FALSE,warning=FALSE}
#FEMALE
FR_female <-life_table_pivot%>%
  dplyr::filter(Country == 'France' , Gender == 'Female')%>%
  dplyr::select(-Country,-Gender)
svdfemaleFR <- svd(FR_female ,nv=2)
summary(svdfemaleFR) %>%
  knitr::kable()
#trouver exposition à la mort  en fct de qx ?
#filtrer life table a usa et selectioner uniquement les dx et qx et age et year 
#esssayer d'avoir une matrice de la meme dimensiion que life table pivot fiotre aux USA 
pivot_life_lx <- function(life_table) {
pivot_life_lx <- life_table %>%
  select(Country, Gender, Year, Age, lx) %>%
    pivot_wider(
      id_cols = c("Country","Gender","Year"),
      names_from = Age,
      values_from = lx
    )
return (pivot_life_lx)
}
l__x <- pivot_life_lx(life_table)
lx_femaleFR<- l__x %>%
  filter(Country == 'France' , Gender == 'Female') %>%
  select(-Country , -Gender )
 
Exp <- life_table$lx
#créer un life table pivot avec lx de mm dim que life_pivot _qx
#pop <- acast(lx,Age~Year)
FR_female <- FR_female%>%
  select( -Year)
lx_femaleFR <- lx_femaleFR%>%
  select( -Year)

FRfemale <- demography::demogdata(data = t(FR_female), 
                 pop = t(lx_femaleFR), 
                 ages = 1:110, 
                 years= 1816:2017, 
                 type = "mortality", 
                 label = "France", 
                 name = "Female",
                 lambda = 1)
LC_female_FR<-demography::lca(FRfemale, max.age = 110)
```

```{r, include=FALSE, echo=FALSE, warning=FALSE}
constLC <- function (ax, bx,kt ,b0x, gc ,wxt ,ages) {
  c1 <- mean(kt[1,], na.rm = TRUE)
  c2 <- sum(bx[, 1], na.rm = TRUE)
  list(ax = ax+c1*bx[,1] ,
       bx=bx[,1]/c2 ,
       kt=c2*(kt[1,]-c1))
}
LC <- StMoMo(link = "log" , staticAgeFun = TRUE , periodAgeFun = "NP" , constFun = constLC)
```
.pull-left[
```{r, echo=TRUE, warning=FALSE}
LC <- lc() #Leecarter function from StMoMo on a demogdata object
FRF <- StMoMoData(FRfemale , series = "Female") #Transforming a demogdata FRfemale object into StMoMoData object
ages.fit <- 1:110
years.fit <- 1945:1995
LCfit1 <- fit(LC, data =FRF , ages.fit=ages.fit , years.fit=years.fit) #Lee Carter Function from StMoMo

#Forecasting 
LCforFRF <- forecast(LCfit1, h=20)
```
]
.pull-right[
```{r, echo=FALSE}
plot(LCforFRF, parametricbx = FALSE)
```
]
]

---
class: center, middle
##III. Estimating life expectancies with demography
.middle[
.pull-left[

```{r,echo=FALSE, warning=FALSE, include=FALSE}
#Getting Ex_pivot_table from life_table

ex_pivot_life <- function(life_table) {
life_table_pivot <- life_table %>%
  select(Country, Gender, Year, Age, ex) %>%
    pivot_wider(
      id_cols = c("Country","Gender","Year"),
      names_from = Age,
      values_from = ex
    )
return (life_table_pivot)
}
ex_life_table_pivot <-ex_pivot_life(life_table)
```
```{r, echo=TRUE, warning=FALSE}
ex_predicted_fr_female <- tibble()
for (a in  0:109){
lf <- flife.expectancy(FRfemale, series = names(FRfemale$rate)[1], years = FRfemale$year,
   age=a+1, max.age = 110)
  lf <- as.vector(t(tibble(lf)))
  ex_predicted_fr_female <- rbind(ex_predicted_fr_female,lf)
}
colnames(ex_predicted_fr_female) <- 1816:2017

ex_predicted_pivot_fr_female <- as.data.frame(t(ex_predicted_fr_female))
Country <- rep("France", 202)
Gender <- rep("Female", 202)
Year <- 1816:2017
colnames(ex_predicted_pivot_fr_female) <- 0:109

ex_predicted_pivot_fr_female <-cbind(Country,Gender, Year,ex_predicted_pivot_fr_female)
ex_observed_pivot_fr_female <- as.data.frame(ex_life_table_pivot %>% filter(Country == "France", Gender == "Female"))
```

```{r, echo=FALSE, warning=FALSE}
ex_observed_fr_female <- ex_observed_pivot_fr_female %>%
  pivot_longer(
      names_to = "Age",
      cols = -c("Year","Country","Gender"),
      values_to =  "Ex")

ex_observed_fr_female <- cbind(rep("Observed",nrow(ex_observed_fr_female)),ex_observed_fr_female)
colnames(ex_observed_fr_female)[1] <-"Type"
colnames(ex_observed_fr_female)[2] <-"Country"


ex_predicted_fr_female <- ex_predicted_pivot_fr_female %>%
  pivot_longer(
      names_to = "Age",
      cols = -c("Year","Country","Gender"),
      values_to =  "Ex")

ex_predicted_fr_female <- cbind(rep("Predicted",nrow(ex_predicted_fr_female)),ex_predicted_fr_female)
colnames(ex_predicted_fr_female)[1] <-"Type"
colnames(ex_predicted_fr_female)[2] <-"Country"
```

```{r,include=FALSE, warning=FALSE, echo=FALSE}
ex_fr_female <- rbind(ex_predicted_fr_female, ex_observed_fr_female)

ex_fr_female %>%
  tibble() %>%
  head()
```
]
.pull-right[
```{r,echo=FALSE, warning=FALSE}


ex_fr_female$Age <- as.numeric(ex_fr_female$Age)
femalefrex<-(ggplot(data = ex_fr_female) +
   aes(x = Year, 
      y = Ex) +
    aes(frame = Age,
        color = ex_fr_female$Type)+
   geom_line() +
  theme(legend.position = "none") +
  labs(   color = "Type" )) %>% 
  plotly::ggplotly(height = 400, width=550 )

femalefrex
```
]
]

